---
title: "Breast cancer vs healthy controls: Elastic net with nested cross-validation"
author: Kat Moore
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    highlight: kate
    df_print: paged
---

An evaluation of baseline elastic net performance using breast cancer and healthy control samples from the TEP dataset using a cross-validation loop.

```{r, include=F}
library(edgeR)
#library(RUVSeq)
#library(openxlsx)
library(RColorBrewer)
library(glmnet)
library(here)
library(foreach)
library(doParallel)
#library(e1071)
library(caret)
library(pROC)
library(tidyverse)
library(ggthemes)
library(ggsci)
library(ggpubr)

theme_set(theme_bw())
```

This will be done using the same dataset we used for the earlier notebooks in this series:

```{r}
dgeFiltered <- readRDS(here("Rds/01_dgeFiltered.Rds")) 
```

This is the TEP dataset we received from the VUMC in Dec 2018.
It has been pre-filtered to select reads which span the splice junction.
In notebook one, we applied a minimum count filter and leave-one-out cross-correlation.
This is the class distribution of samples that remain:

```{r}
table(dgeFiltered$samples$group)
```

## Leave-one-out cross-validation

We ran LOOCV in `05_enet_xval.R`. It took around 3 hours to complete with 16 cores.

```{r load results}
modelList <- readRDS(here("Rds", "05_LOOCV_enet.Rds"))
```

We should have as many models as we do samples.

```{r}
length(modelList) == ncol(dgeFiltered)
```
As before, each element in the list has three components: fit, train and test.

Fit is a train-class object produced by caret, with alpha and lambda tuned.
Train and test are both lists containing elements matching either the training samples or the left-out sample (test).

The lists contain:
a) data, a samples x genes matrix used to either train the data or representing the left-out sample
b) labels, the true class labels belonging to the samples in that data partition.

For example:

```{r}
modelList[[1]]$fit$bestTune
modelList[[1]]$train$data[1:4,1:4]
modelList[[1]]$test$labels
```

Apply names to list to match the left out sample:

```{r}
names(modelList) <- lapply(
  modelList, function(x) x$test$labels$sample_name) %>%
  unlist()

stopifnot(names(modelList)[[1]] == modelList[[1]]$test$labels$sample_name)
```

### Hyperparameters

The majority of models selected alpha = 0.1.

```{r}
hyperparams <- lapply(modelList, function(x) x$fit$bestTune) %>%
  bind_rows(.id = "sample") 

hyperparams %>%
  group_by(alpha) %>%
  count()
```

In a few cases where alpha = 0, lambda is very high.

```{r}
hyperparams %>%
  ggplot(aes(x = alpha, y = lambda)) +
  geom_jitter() +
  scale_y_log10() +
  ggtitle("Best alpha and lambda in LOOCV")
```

## Predictions

Predict the left-out sample for the entire list.

```{r make predictions}
#modelList[[1]]$test$data[,1:4, drop=F]

enet_pred <- function(x){
  
  fit <- x$fit
  val_data <- x$test$data
  
  class = cbind(
    sample = rownames(val_data),
    predicted.group = as.character(predict(fit, newdata = val_data, type="raw"))
  )
  
  #return(class)
  
  probs = predict(fit, newdata = val_data, type="prob")
  #return(probs)
  
  colnames(probs) <- paste0("prob.",colnames(probs))
  probs$model <- "Enet LOOCV"
  
  
  
  cbind(class, probs)  
  
}

#enet_pred(modelList[[1]])

predictions <- lapply(modelList, enet_pred) %>%
  bind_rows()

head(predictions)

#Should be no NAs
stopifnot(all(rowSums(is.na(predictions))==0))
```

Add true class labels so we can assess performance, alongside stage and hyperparameters.

```{r}
predictions <- dgeFiltered$samples %>%
  rownames_to_column("sample") %>%
  select(sample, true.group = group, stage, isolationlocation) %>%
  right_join(., predictions, by = "sample") %>%
  left_join(., hyperparams, by = "sample")

#Ensure levels are in the same order
predictions$true.group <- factor(predictions$true.group, levels=c("healthyControl", "breastCancer"))
predictions$predicted.group <- factor(predictions$predicted.group, levels=c("healthyControl", "breastCancer"))

head(predictions)
```

## Performance

```{r}
predictions <- predictions %>% 
  mutate(misclassified = (predicted.group != true.group))
```

### For all stages:

```{r}
auc_all <- pROC::roc(response = predictions$true.group,
                     predictor = predictions$prob.breastCancer)

auc_all
```

CI:

```{r}
pROC::ci.auc(auc_all)
```

#### ROC

```{r aucplot1}
plotmyroc <- function(roc.res, title){
  rocobj <- pROC::plot.roc(roc.res,
                           main = title, 
                           percent=TRUE,
                           #ci = TRUE,  #Finite xlim glitch         
                           print.auc = TRUE,
                           asp = NA)
  
  ciobj <- ci.se(rocobj,                         # CI of sensitivity
                 specificities = seq(0, 1, 0.05)) # over a select set of specificities
  plot(ciobj, type = "shape", col = "#1c61b6AA")     # plot as a blue shape
  plot(ci(rocobj, of = "thresholds", thresholds = "best")) # add one threshold
  
}


plotmyroc(auc_all, title = "Elastic net LOOCV: all stages")
```

Number of samples misclassified. Unsurprisingly, we miss more in the early stage.

```{r}
table(predictions$misclassified, predictions$stage)
```

#### Boxplot

```{r}
predictions %>%
  ggplot(aes(x = true.group, y = prob.breastCancer)) +
  geom_jitter(aes(color = isolationlocation), height = 0, width = 0.1) +
  geom_boxplot(alpha = 0) +
  ggtitle("LOOCV enet predictions: all stages")
```

### Early stage (I and II)

```{r}
predictions_early <- filter(predictions, stage %in% c("I", "II", "healthyControl"))

auc_early <- pROC::roc(response = predictions_early$true.group,
                       predictor = predictions_early$prob.breastCancer)

auc_early
```

CI:

```{r}
pROC::ci.auc(auc_early)
```

#### ROC

```{r aucplot2}
plotmyroc <- function(roc.res, title){
  rocobj <- pROC::plot.roc(roc.res,
                           main = title, 
                           percent=TRUE,
                           #ci = TRUE,  #Finite xlim glitch         
                           print.auc = TRUE,
                           asp = NA)
  
  ciobj <- ci.se(rocobj,                         # CI of sensitivity
                 specificities = seq(0, 1, 0.05)) # over a select set of specificities
  plot(ciobj, type = "shape", col = "#1c61b6AA")     # plot as a blue shape
  plot(ci(rocobj, of = "thresholds", thresholds = "best")) # add one threshold
  
}


plotmyroc(auc_early, title = "Elastic net LOOCV: early stage (I and II)")
```

Number of samples misclassified:

```{r}
table(predictions_early$misclassified, predictions_early$stage)
```

#### Boxplot

```{r}
predictions_early %>%
  ggplot(aes(x = true.group, y = prob.breastCancer)) +
  geom_jitter(aes(color = isolationlocation), height = 0, width = 0.1) +
  geom_boxplot(alpha = 0) +
  ggtitle("LOOCV enet predictions: early stages")
```

### Late stage (III and IV)

```{r}
predictions_late <- filter(predictions, stage %in% c("III", "IV", "healthyControl"))

auc_late <- pROC::roc(response = predictions_late$true.group,
                       predictor = predictions_late$prob.breastCancer)

auc_late
```

CI:

```{r}
pROC::ci.auc(auc_late)
```

```{r aucplot3}
plotmyroc <- function(roc.res, title){
  rocobj <- pROC::plot.roc(roc.res,
                           main = title, 
                           percent=TRUE,
                           #ci = TRUE,  #Finite xlim glitch         
                           print.auc = TRUE,
                           asp = NA)
  
  ciobj <- ci.se(rocobj,                         # CI of sensitivity
                 specificities = seq(0, 1, 0.05)) # over a select set of specificities
  plot(ciobj, type = "shape", col = "#1c61b6AA")     # plot as a blue shape
  plot(ci(rocobj, of = "thresholds", thresholds = "best")) # add one threshold
  
}


plotmyroc(auc_late, title = "Elastic net LOOCV: late stage (III and IV)")
```

Number of samples misclassified:

```{r}
table(predictions_late$misclassified, predictions_late$stage)
```

## Confusion matrices

All stages:

```{r}
confMat <- caret::confusionMatrix(data = predictions$predicted.group,
                                  reference = predictions$true.group,
                                  positive = "breastCancer",
                                  mode = "everything")

confMat
```

Early stage:

```{r}
confMat_early <- caret::confusionMatrix(data = predictions_early$predicted.group,
                                        reference = predictions_early$true.group,
                                        positive = "breastCancer",
                                        mode = "everything")

confMat_early
```

Late stage:

```{r}
confMat_late <- caret::confusionMatrix(data = predictions_late$predicted.group,
                                        reference = predictions_late$true.group,
                                        positive = "breastCancer",
                                        mode = "everything")

confMat_late
```

## Save results

```{r}
pretty_ci <- function(auc, conf.level = 0.95, return){
  stopifnot(return %in% c("lower","upper","range"))
  
  if(return == "lower"){
    return(signif(pROC::ci(auc, conf.level = conf.level)[1], 4))
  }
  
  if(return == "upper"){
    return(signif(pROC::ci(auc, conf.level = conf.level)[3], 4))
  }
  
  if(return == "range"){
    paste(signif(pROC::ci(auc, conf.level = conf.level)[1], 4),
        signif(pROC::ci(auc, conf.level = conf.level)[3], 4),
        sep="-")
  }
  
}


aucs <- list(auc_all, auc_early, auc_late)
confMats <- list(confMat, confMat_early, confMat_late)


auc.summary <- tibble(
  Stage = c("All stages", "Early (I-II)", "Late (III-IV)"),
  AUC = sapply(aucs, function(x) x$auc),
  CI.95.lower = sapply(aucs, function(x) pretty_ci(x, return = "lower")),
  CI.95.upper = sapply(aucs, function(x) pretty_ci(x, return = "upper")),
  Sensitivity = sapply(confMats, function(x) x$byClass[names(x$byClass)=="Sensitivity"]),
  Specificity = sapply(confMats, function(x) x$byClass[names(x$byClass)=="Specificity"]),
  Precision = sapply(confMats, function(x) x$byClass[names(x$byClass)=="Precision"]),
  Recall = sapply(confMats, function(x) x$byClass[names(x$byClass)=="Recall"]),
  F1 = sapply(confMats, function(x) x$byClass[names(x$byClass)=="F1"])
  )


auc.summary
```

Save to csv

```{r}
write_csv(auc.summary, here("05_enet_LOOCV_performance.csv"))
write_csv(predictions, here("05_enet_LOOCV_predictions.csv"))
```

```{r}
sessionInfo()
```
