---
title: "Post hoc analyses: dimensionality"
author: "Kat Moore"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 5
    df_print: paged
    highlight: kate
---

```{r setup, include=FALSE}
library(edgeR)
library(vsn)
library(DESeq2)
library(here)
library(caret)
library(ComplexHeatmap)
library(Rtsne)
library(umap)
library(pROC)
library(tidyverse)
library(ggthemes)
library(ggsci)

theme_set(theme_bw())
```

The performance of all TEP-based classifiers on the blind validation set was much poorer than expected.
In this notebook, we will explore possible explanations for the poor performance, focusing on dimensionality analyses. 

## Load data

Expression & metadata for both original dataset and blind validation:

```{r}
dgeAll <- readRDS(file = here("Rds/07b_dgeAll.Rds"))
```

Create subsets for original data and blind val:

```{r}
dgeOriginal <- dgeAll[,dgeAll$samples$Dataset == "Original"]
dgeOriginal$samples <- droplevels(dgeOriginal$samples)

dgeBlindVal <- dgeAll[,dgeAll$samples$Dataset == "blindVal"]
dgeBlindVal$samples <- droplevels(dgeBlindVal$samples)
```

Elastic net model, from which we extract the features

```{r}
model_altlambda <- readRDS(here("Rds/04_model_altlambda.Rds"))

feature_extraction <- function(fit, dict = dgeAll$genes){
   coef(fit$finalModel, fit$bestTune$lambda) %>%
    as.matrix() %>% as.data.frame() %>%
    slice(-1) %>% #Remove intercept
    rownames_to_column("feature") %>%
    rename(coef = "1") %>%
    filter(coef !=0) %>%
    left_join(., dict, by = c("feature" = "ensembl_gene_id")) %>%
    rename("ensembl_gene_id"=feature) %>%
    arrange(desc(abs(coef)))
}

enet_feat <- feature_extraction(model_altlambda$fit)
```

Total number of elastic net features in the final version:

```{r}
enet_feat %>% nrow()
```

PSO-SVM model & features:

```{r}
pso_output <- readRDS(here("Rds/02_thromboPSO.Rds"))

best.selection.pso <- paste(pso_output$lib.size,
                            pso_output$fdr,
                            pso_output$correlatedTranscripts,
                            pso_output$rankedTranscripts, sep="-")

particle_path <- file.path(here("pso-enhanced-thromboSeq1/outputPSO", 
                           paste0(best.selection.pso,".RData")))

load(particle_path) #Becomes dgeTraining
dgeParticle <- dgeTraining #Rename to avoid namespace confusion
rm(dgeTraining)


#Features from PSO-SVM do not have coefficients, 
#but they should come out of Thromboseq in a ranked order.

psosvm.feat <- enframe(dgeParticle$biomarker.transcripts, "rank", "ensembl_gene_id") %>%
  left_join(dgeParticle$genes, by = "ensembl_gene_id")
```

Number of PSO-SVM features:

```{r}
psosvm.feat %>% nrow()
```

## PCA

For PCA we want to render the data homoskedastic: that is, the variances should be mostly the same and not dependent on large counts.
We have three possible ways of doing this:

a) voom transformation
b) log-transformed TMM counts
c) variance stabilizing transformation

```{r}
v <- limma::voom(dgeAll, design = model.matrix(~group, dgeAll$samples), plot = F)
normTMMlog2 <- function(object){
  object = calcNormFactors(object, method="TMM")
  object = cpm(object, log=T, normalized.lib.sizes=T)
  return(object)
}

normAll <- normTMMlog2(dgeAll)
vsdAll <- DESeqDataSetFromMatrix(countData = dgeAll$counts,
                                 colData = dgeAll$samples,
                                 design = ~1) %>%
  vst(., blind=T)
```

### Mean-sd plots

Plot the mean by the standard deviation. We want to see as straight a line as possible.

```{r}
#Need to set plot=F and then access gg to add a title
vsn::meanSdPlot(v$E, plot=F)$gg + ggtitle("Voom")
vsn::meanSdPlot(normAll, plot=F)$gg + ggtitle("Log TMM")
vsn::meanSdPlot(assay(vsdAll), plot=F)$gg + ggtitle("Vst")
```

Of the three, voom is clearly the worst. TMM and VST are pretty close. We'll go with TMM, because the rest of the classifiers have been trained on TMM-normalized data, and the dimensionality analyses resulting from a TMM-based strategy will be closest to what the classifier "sees".

Note: in the previous analysis, we used VST instead of TMM. As such, these visualizations will look a little different than they do in `pso-vs-hc`.

```{r, include=F}
#Remove the unused objects:
rm(vsdAll)
rm(v)
```

Normalize the subsetted objects:

```{r}
normBlind <- normTMMlog2(dgeBlindVal)
normOriginal <- normTMMlog2(dgeOriginal)
```

Set up some functions for computing & plotting PCAs.

```{r}
#Will return a data frame instead of a plot if returnData is TRUE
scree_plot = function(res_prcomp = nc.pca, n_pca=20, returnData = F){
  
  percentVar <- res_prcomp$sdev^2/sum(res_prcomp$sdev^2) * 100
  names(percentVar) = paste0("PC", seq(1, length(percentVar)))
  percentVar = enframe(percentVar, "PC", "variance")
  
  percentVar$PC=factor(percentVar$PC, levels=percentVar$PC)
  if(returnData){return(percentVar)}
  percentVar %>%
    dplyr::slice(1:n_pca) %>%
    ggplot(aes(x=PC, y=variance)) +
    geom_bar(stat="identity") +
    ggtitle("Scree plot") +
    ylab("Percent total variance") +
    theme_bw()
  
}
```

```{r}
#For extracting percent variance associated with a PC
get_var <- function(PC, x){
  p = x[x$PC == PC, ]$variance
  paste0(round(p, 1), "%")
}
```

```{r}
#Plot PCAs with ggplots2, color mandatory, shape optional
gg_pca <- function(pca.df, x.pc, y.pc, pVar,
                   color.var, shape.var=NULL){
  
  ncolors <- length(unique(pca.df[,color.var]))
  
  if(is.null(shape.var)){
    ggplot(pca.df, aes(x=get(x.pc), y=get(y.pc),
                       color=get(color.var))) +
      geom_point()  + 
      xlab(paste0(x.pc, ": ", get_var(x.pc,x = pVar))) +
      ylab(paste(y.pc, ": ", get_var(y.pc,x = pVar))) +
      labs(color = color.var)
  } else {
    ggplot(pca.df, aes(x=get(x.pc), y=get(y.pc),
                       color=get(color.var),
                       shape=get(shape.var))) +
      geom_point()  +
      xlab(paste0(x.pc, ": ", get_var(PC = x.pc, x = pVar))) +
      ylab(paste(y.pc, ": ", get_var(y.pc,x = pVar))) +
      labs(color = color.var, shape = shape.var)
  }
}
```

```{r}
#Wrapper function to go from a count matrix straight to a ggplot of PCA data
mat_to_ggpca <- function(mat, cvar = "group", svar = NULL,
                         features = NULL, # an optional vector of features
                         sampledata,
                         pc.x = "PC1", pc.y = "PC2", returnData = F){
  
  #Select relevant features
  if(!is.null(features)){
    stopifnot(all(features %in% rownames(mat)))
    mat <- mat[rownames(mat) %in% features,] 
    stopifnot(nrow(mat) == nrow(features))
  }
  
  #Calculate PCs
  thisnc.pca <- prcomp(t(mat))
  thispercVar <- scree_plot(res_prcomp = thisnc.pca, n_pca=20, returnData = T)
  
  #Sometimes less than 100 PCs are returned
  pcmax <- ifelse(ncol(thisnc.pca$x) < 100, ncol(thisnc.pca$x), 100)
  
  #Reshape data
  thispca.df <- as.data.frame(thisnc.pca$x[, 1:pcmax]) %>%
    rownames_to_column("sample")
  thispca.df <- right_join(sampledata,thispca.df, by='sample')
  
  if(returnData){return(thispca.df)}
  
  #Create plot
  gg_pca(pca.df = thispca.df, x.pc = pc.x, y.pc = pc.y, pVar = thispercVar,
       color.var = cvar, shape.var = svar)
}

```

### Scree plot

```{r}
scree_plot(prcomp(t(normAll)))
```

### All samples

#### Case-control status

```{r}
mat_to_ggpca(normAll, cvar = "group", sampledata = dgeAll$samples)+
  scale_color_calc() +
  ggtitle("Case control status: all samples")
```

#### Elastic net features

```{r}
mat_to_ggpca(normAll, features = enet_feat$ensembl_gene_id,
             sampledata = dgeAll$samples, cvar = "group") +
  scale_color_calc() +
  ggtitle(paste("PCA on", length(enet_feat$ensembl_gene_id),
                "enet features, all samples"))
```

#### PSO features

```{r}
mat_to_ggpca(normAll, features = psosvm.feat$ensembl_gene_id,
             sampledata = dgeAll$samples, cvar = "group") +
  scale_color_calc() +
  ggtitle(paste("PCA on", length(psosvm.feat$ensembl_gene_id),
                "pso-svm features, all samples"))
```

#### Hospital

```{r}
mat_to_ggpca(normAll, cvar = "hosp", sampledata = dgeAll$samples)+
  scale_color_aaas() +
  ggtitle("Hospital of origin: all samples")

```

### Original dataset

#### Case-control status

```{r}
mat_to_ggpca(normOriginal, cvar = "group", sampledata = dgeOriginal$samples)+
  scale_color_calc() +
  ggtitle("Case control status: original dataset")
```

#### Elastic net features

```{r}
mat_to_ggpca(normOriginal, features = enet_feat$ensembl_gene_id,
             sampledata = dgeOriginal$samples, cvar = "group") +
  scale_color_calc() +
  ggtitle(paste("PCA on", length(enet_feat$ensembl_gene_id),
                "enet features, original dataset"))
```

Subsequent PCs:

```{r}
mat_to_ggpca(normOriginal, features = enet_feat$ensembl_gene_id,
             cvar = "group", sampledata = dgeOriginal$samples, pc.x = "PC3", pc.y = "PC4") +
  scale_color_calc() +
  ggtitle(paste("PCA on", nrow(enet_feat), "enet features, original dataset"))

mat_to_ggpca(normOriginal, features = enet_feat$ensembl_gene_id,
             cvar = "group", sampledata = dgeOriginal$samples, pc.x = "PC5", pc.y = "PC6") +
  scale_color_calc() +
  ggtitle(paste("PCA on", nrow(enet_feat), "enet features, original dataset"))

mat_to_ggpca(normOriginal, features = enet_feat$ensembl_gene_id,
             cvar = "group", sampledata = dgeOriginal$samples, pc.x = "PC7", pc.y = "PC8") +
  scale_color_calc() +
  ggtitle(paste("PCA on", nrow(enet_feat), "enet features, original dataset"))
```

PCs 6-8 look a little better than PCs 1&2, but not a lot. Can we do better?

##### PCs associated with case-control status

Testing the first 100 principal components for significance with cancer/noncancer.

```{r}
#Combine metadata with PCs
pca.df <- mat_to_ggpca(normOriginal, features = enet_feat$ensembl_gene_id,
                       sampledata = dgeOriginal$samples, cvar = "group", returnData = T)

#Perform kruskal wallis test and extract p value
pc_cancer <- lapply(paste0('PC', 1:100), function(x){
  kruskal.test(pca.df[[x]], pca.df$group)$p.value
}) %>%
  #Reshape results
  unlist() %>% enframe("PC", "nom.p") %>%
  #Multiple testing correction
  mutate(bonferroni=p.adjust(nom.p, "bonferroni"),
         fdr = p.adjust(nom.p, "BH")) %>%
  #Select significant PCs and order by fdr
  filter(fdr < 0.05) %>% arrange(fdr) #%>%
  #mutate(PCs = str_pad(as.character(PC), width = 2, side = "left", pad = "0"), .after= PC)

pc_cancer %>%
  arrange(PC)
```

Variance represented by all PCs significantly associated with case control status in original dataset. Note that this is not the same as variance explained by case control status.

```{r}
prcomp(t(normOriginal)) %>%
  scree_plot(res_prcomp = ., n_pca=100, returnData = T) %>%
  mutate(PC = as.integer(str_remove(as.character(PC), "PC"))) %>%
  filter(PC %in% pc_cancer$PC) %>% pull(variance) %>%
  sum()
```

Plot PCs associated with case-control status.

```{r}
mat_to_ggpca(normOriginal, features = enet_feat$ensembl_gene_id,
             sampledata = dgeOriginal$samples,
             cvar = "group", pc.x = "PC1", pc.y = "PC6") +
  scale_color_calc() +
  ggtitle(paste("Top PCs (1 & 6) significantly associated with case control status,
                elastic net features, original dataset"))
```

```{r}
mat_to_ggpca(normOriginal, features = enet_feat$ensembl_gene_id,
             sampledata = dgeOriginal$samples,
             cvar = "group", pc.x = "PC6", pc.y = "PC7") +
  scale_color_calc() +
  ggtitle(paste("PCs (6 & 7) significantly associated with case control status,
                elastic net features, original dataset"))
```

#### PSO features

```{r}
mat_to_ggpca(normOriginal, features = psosvm.feat$ensembl_gene_id,
             sampledata = dgeOriginal$samples, cvar = "group") +
  scale_color_calc() +
  ggtitle(paste("PCA on", length(psosvm.feat$ensembl_gene_id),
                "pso-svm features, original dataset"))
```

#### Hospital

```{r}
mat_to_ggpca(normOriginal, cvar = "hosp", sampledata = dgeOriginal$samples)+
  scale_color_aaas() +
  ggtitle("Hospital of origin: original dataset")

```

### Blind validation dataset

#### Case-control status

```{r}
mat_to_ggpca(normBlind, cvar = "group", sampledata = dgeBlindVal$samples)+
  scale_color_calc() +
  ggtitle("Case control status: blind validation")
```

#### Elastic net features

```{r}
mat_to_ggpca(normBlind, features = enet_feat$ensembl_gene_id,
             sampledata = dgeBlindVal$samples, cvar = "group") +
  scale_color_calc() +
  ggtitle(paste("PCA on", length(enet_feat$ensembl_gene_id),
                "enet features, blind validation dataset"))
```

#### PSO features

```{r}
mat_to_ggpca(normBlind, features = psosvm.feat$ensembl_gene_id,
             sampledata = dgeBlindVal$samples, cvar = "group") +
  scale_color_calc() +
  ggtitle(paste("PCA on", length(psosvm.feat$ensembl_gene_id),
                "pso features, blind validation dataset"))
```

#### Ery contamination

We only have this metric available for some of the blind validation samples, scored by VUMC techs.

```{r}
mat_to_ggpca(normBlind,
             sampledata = dgeBlindVal$samples,
             cvar = "ery_contam") +
  scale_color_calc() +
  ggtitle(paste("PCA on blind validation dataset: ery contamination"))
```

#### Pellet quality

As with ery contam, we only have this metric available for some of the blind validation samples, scored by VUMC techs.

```{r}
mat_to_ggpca(normBlind,
             sampledata = dgeBlindVal$samples,
             cvar = "pellet_qual") +
  scale_color_calc() +
  ggtitle(paste("PCA on blind validation dataset: pellet quality"))
```

## t-SNE

### Case control status

On original dataset:

```{r}
#set.seed(123)

gg_tsne <- function(mat, sampledata, col = "group", seed = 123, returndata = F){
  
  #A seed must be set for reproducible results
  set.seed(seed)
 
  #Input should be row observations x column variables
  tsne <- Rtsne::Rtsne(t(mat))
  
  tsne_df <- data.frame(tsne_x = tsne$Y[,1], tsne_y = tsne$Y[,2])
    
  tsne_df <- bind_cols(tsne_df, sampledata)
  
  if(returndata){return(tsne_df)}
  
  tsne_df %>%
    ggplot(aes(x = tsne_x, y = tsne_y, color = get(col))) +
    geom_point() +
    labs(color = col)
}

gg_tsne(mat=normOriginal,
        sampledata = dgeOriginal$samples) +
  ggthemes::scale_color_calc() +
  ggtitle("t-SNE on original dataset, case-control status")

```

Facet by hospital:

```{r}
gg_tsne(mat=normOriginal,
        sampledata = dgeOriginal$samples) +
  ggthemes::scale_color_calc() +
  ggtitle("t-SNE on original dataset, case-control status") +
  facet_wrap(~hosp)
```

### Hospital of origin

```{r}
gg_tsne(mat=normOriginal,
        sampledata = dgeOriginal$samples,
        col = "hosp") +
  ggsci::scale_color_aaas() +
  ggtitle("t-SNE on original dataset, isolation location")
```

### EN features only

Subset counts to genes present within elastic net.

```{r}
normEN <- normOriginal[rownames(normOriginal) %in% enet_feat$ensembl_gene_id, ]
stopifnot(nrow(normEN) == nrow(enet_feat))
```

```{r}
gg_tsne(mat=normEN,
        sampledata = dgeOriginal$samples) +
  ggthemes::scale_color_calc() +
  ggtitle("t-SNE on original dataset, elastic net features")
```

```{r}
gg_tsne(mat=normEN,
        sampledata = dgeOriginal$samples) +
  ggthemes::scale_color_calc() +
  ggtitle("t-SNE on original dataset, elastic net features") +
  facet_wrap(~hosp)
```

```{r}
gg_tsne(mat=normEN, col = "hosp",
        sampledata = dgeOriginal$samples) +
  ggsci::scale_color_aaas() +
  ggtitle("t-SNE on original dataset, elastic net features")
```

## UMAP

See [vignette](https://cran.r-project.org/web/packages/umap/vignettes/umap.html) for details.

### Case-control status

```{r}
gg_umap <- function(mat, sampledata, col = "group", seed = 123, returndata = F){
  
  #Set seed for reproducibility
  set.seed(seed)
  
  #input should be observations x variables
  mapres <- umap(t(mat))
  
  #Create data frame from umap results and add sample data
  umap_df <- data.frame(umap_x = mapres$layout[,1], umap_y = mapres$layout[,2])
  umap_df <- bind_cols(umap_df, sampledata)
  
  if(returndata){return(umap_df)}
  
  #Plot
  umap_df %>%
    ggplot(aes(x = umap_x, y = umap_y, color = get(col))) +
    geom_point() +
    labs(color = col)
  
}

gg_umap(mat=normOriginal,
        sampledata = dgeOriginal$samples,
        col = "group") +
  ggthemes::scale_color_calc() +
  ggtitle("UMAP on original dataset, case-control status")
```

Faceted by hospital.

```{r}
gg_umap(mat=normOriginal,
        sampledata = dgeOriginal$samples,
        col = "group") +
  ggthemes::scale_color_calc() +
  ggtitle("UMAP on original dataset, case-control status") +
  facet_wrap(~hosp)
```

### Hospital of origin

```{r}
gg_umap(mat=normOriginal,
        sampledata = dgeOriginal$samples,
        col = "hosp") +
  ggsci::scale_color_aaas() +
  ggtitle("UMAP on original dataset, isolation location")
```

```{r}
sessionInfo()
```
