---
title: "Feature plots"
author: "Kat Moore"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 5
    df_print: paged
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(edgeR)
library(here)
library(ComplexHeatmap)
library(tidyverse)
library(ggthemes)
library(ggsci)

theme_set(theme_bw())
```

Plot expression of top ranked elastic net and PSO-SVM features as box plots or as heatmaps. 

## Load data

Expression & metadata for both original dataset and blind validation:

```{r}
dgeAll <- readRDS(file = here("Rds/07b_dgeAll.Rds"))
```

Create subsets for original data and blind val:

```{r}
dgeOriginal <- dgeAll[,dgeAll$samples$Dataset == "Original"]
dgeOriginal$samples <- droplevels(dgeOriginal$samples)

dgeBlindVal <- dgeAll[,dgeAll$samples$Dataset == "blindVal"]
dgeBlindVal$samples <- droplevels(dgeBlindVal$samples)
```

Elastic net model, from which we extract the features

```{r}
model_altlambda <- readRDS(here("Rds/04_model_altlambda.Rds"))

feature_extraction <- function(fit, dict = dgeAll$genes){
   coef(fit$finalModel, fit$bestTune$lambda) %>%
    as.matrix() %>% as.data.frame() %>%
    slice(-1) %>% #Remove intercept
    rownames_to_column("feature") %>%
    rename(coef = "1") %>%
    filter(coef !=0) %>%
    left_join(., dict, by = c("feature" = "ensembl_gene_id")) %>%
    rename("ensembl_gene_id"=feature) %>%
    arrange(desc(abs(coef))) %>%
    rowid_to_column("rank")
}

enet_feat <- feature_extraction(model_altlambda$fit)

head(enet_feat)
```

Total number of elastic net features in the final version:

```{r}
enet_feat %>% nrow()
```

PSO-SVM model & features:

```{r}
pso_output <- readRDS(here("Rds/02_thromboPSO.Rds"))

best.selection.pso <- paste(pso_output$lib.size,
                            pso_output$fdr,
                            pso_output$correlatedTranscripts,
                            pso_output$rankedTranscripts, sep="-")

particle_path <- file.path(here("pso-enhanced-thromboSeq1/outputPSO", 
                           paste0(best.selection.pso,".RData")))

load(particle_path) #Becomes dgeTraining
dgeParticle <- dgeTraining #Rename to avoid namespace confusion
rm(dgeTraining)


#Features from PSO-SVM do not have coefficients, 
#but they should come out of Thromboseq in a ranked order.

psosvm.feat <- enframe(dgeParticle$biomarker.transcripts, "rank", "ensembl_gene_id") %>%
  left_join(dgeParticle$genes, by = "ensembl_gene_id")

head(psosvm.feat)
```

Number of PSO-SVM features:

```{r}
psosvm.feat %>% nrow()
```

## Overview

```{r}
enet_feat %>% head(30)
```

```{r}
psosvm.feat %>% head(30)
```

## Boxplots

### Elastic net features

```{r}
ggfeaturePlot <- function(dge, ens, df){
  
  stopifnot(length(ens) == 1)
  dge$samples$sample <- colnames(dge)
  
  #Normalize
  counts <- cpm(calcNormFactors(dge), log = T, normalized.lib.sizes = T)
  
  #Select relevant gene, convert to tidy format
  counts <- as.data.frame(counts) %>%
    rownames_to_column("ensembl_gene_id") %>%
    filter(ensembl_gene_id == ens) %>%
    select(-ensembl_gene_id) %>%
    gather(key = "sample", value = "exp")
  
  #Add metadata
  counts <- left_join(counts, dge$samples, by = "sample")
  
  gndf <- df %>% filter(ensembl_gene_id == ens)
  
  gn <- ifelse(gndf$hgnc_symbol == "", ensembl_gene_id,
               paste(gndf$hgnc_symbol, gndf$ensembl_gene_id, sep=":"))
  
  title <- paste0(gn, ", Rank: ", gndf$rank)
  
  #Add coef if plotting enet
  if("coef" %in% colnames(gndf)){
    title <- paste0(title, ", Coef: ", signif(gndf$coef, 3))
  }
  
  counts %>%
    ggplot(aes(x = group, y = exp)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(aes(color = hosp), alpha = 0.7, height = 0, width = 0.25) +
    ggsci::scale_color_igv() +
    ggtitle(title) +
    facet_wrap(~Dataset)
  
}

#ggfeaturePlot(dgeAll, ens = "ENSG00000113269", enet_feat)

lapply(enet_feat$ensembl_gene_id[1:10],
       function(x) ggfeaturePlot(dgeAll, ens = x, enet_feat))

```

### PSO-SVM features

```{r}
lapply(psosvm.feat$ensembl_gene_id[1:10],
       function(x) ggfeaturePlot(dgeAll, ens = x, psosvm.feat))

```

## Heatmaps

Plot elastic net features as a heatmap.

### EN features in original data

```{r}
#Heatmap plotting
feature_heatmap = function(dge,
                           title="Heatmap", debug = F,
                           top_vars = c("group","hosp"),
                           features,
                           legend_title = NULL,
                           row_scale = F,
                           row_size = 8, col_size = 8,
                           show_col_names = F, show_row_names = T, ...){
  
  #Define colors
  hmap_colors = list(
    group = c(healthyControl="lightgray", breastCancer="black"),
    hosp = ggsci::pal_lancet()(length(unique(dge$samples$hosp))),
    Dataset = ggsci::pal_jco()(length(unique(dge$samples$Dataset)))
  )
  names(hmap_colors$hosp) = unique(dge$samples$hosp)
  names(hmap_colors$Dataset) = unique(dge$samples$Dataset)
  
  top_colors = list(
    #Dataset = hmap_colors$Dataset,
    group = hmap_colors$group,
    hosp = hmap_colors$hosp
    )
  #bottom_colors = list(group = hmap_colors$group)
  
  #Subset to only those genes used as features
  dge <- dge[rownames(dge) %in% features, ]
  
  #Normalized count matrix
  mat = edgeR::cpm(edgeR::calcNormFactors(dge),
                   log = T, normalized.lib.sizes = T)
  
  #Row scale settings
  if (row_scale==T){
    mat = t(scale(t(mat)))
  }
  #Change legend according to whether input is scaled
  if (row_scale==T){
    hlp = list(title="rowscaled logcpm")
  } else {
    hlp = list(title="logcpm counts")
  }
  
  #Replace ensembl IDs with gene names, or keep ensembl ID if gene names are absent
  genes <- dge$genes
  genes <- genes %>%
    dplyr::mutate(gene_name = as.character(hgnc_symbol)) %>%
    dplyr::mutate(
      label = ifelse(gene_name == "" | is.na(gene_name),
                    ensembl_gene_id, gene_name)
      )
  
  #return(genes)
  rownames(mat) <- genes$label
  
  sampledata <- as.data.frame(dge$samples)
  
  #Heatmap annotation
  ann_top = sampledata[,top_vars, drop=F]
  
  #Top column annotation
  colTop <- ComplexHeatmap::HeatmapAnnotation(
    df=ann_top, which="col",
    col = top_colors
    #annotation_legend_param = list(list(title = legend_title))
  )
  #return(colTop)
  
  #Bottom column annotation
  # ann_bottom = sampledata[,bottom_vars, drop=F]
  # colBottom <- ComplexHeatmap::HeatmapAnnotation(
  #   df=ann_bottom, which="col", col = bottom_colors
  #   )
  
  if(debug){print(dim(mat))}
  stopifnot(nrow(mat) == nrow(features))
  
  #Draw the heatmap
  ComplexHeatmap::Heatmap(mat,
          top_annotation = colTop,
          #bottom_annotation = colBottom,
          #left_annotation = rowAnno,
          heatmap_legend_param = hlp,
          show_row_names = show_row_names,
          show_column_names = show_col_names,
          cluster_rows = T,
          row_names_gp = gpar(fontsize = row_size),
          column_names_gp = gpar(fontsize = col_size),
          column_title = title,
          ...)
}


suppressMessages(
  feature_heatmap(dgeOriginal,
                   features = enet_feat$ensembl_gene_id,
                  show_row_names = F,
                  title = paste0("Elastic net features (", nrow(enet_feat), ") in original data"),
                  row_scale = T)
)

```

### EN features in blind validation data

```{r}
suppressMessages(
  feature_heatmap(dgeBlindVal,
                  features = enet_feat$ensembl_gene_id,
                  show_row_names = F,
                  title = paste0("Elastic net features (", nrow(enet_feat), ") in blind validation data"),
                  row_scale = T)
)
```

### Enet features in all data

```{r}
suppressMessages(
  feature_heatmap(dgeAll,
                  features = enet_feat$ensembl_gene_id,
                  show_row_names = F,
                  title = paste0("Elastic net features (", nrow(enet_feat), ") in all data"),
                  row_scale = T)
)
```

### PSO-SVM features in original data

```{r}
suppressMessages(
  feature_heatmap(dgeOriginal,
                  features = psosvm.feat$ensembl_gene_id,
                  show_row_names = F,
                  title = paste0("PSO-SVM features (", nrow(psosvm.feat), ") in original data"),
                  row_scale = T)
)
```

### PSO-SVM features in blind data

```{r}
suppressMessages(
  feature_heatmap(dgeBlindVal,
                  features = psosvm.feat$ensembl_gene_id,
                  show_row_names = F,
                  title = paste0("PSO-SVM features (", nrow(psosvm.feat), ") in blind data"),
                  row_scale = T)
)
```

### PSO-SVM features in all data

```{r}
suppressMessages(
  feature_heatmap(dgeAll,
                  features = psosvm.feat$ensembl_gene_id,
                  show_row_names = F,
                  title = paste0("PSO-SVM features (", nrow(psosvm.feat), ") in original data"),
                  row_scale = T)
)
```

```{r}
sessionInfo()
```
